### 2016-07-26 report/paper_draft.tex
finished comparison section, with minor changes elsewhere

The goal of this paper is to survey the landscape of deep learning frameworks with support for distributed execution, and compare them according to a consistent set of characteristics. The characteristics compared across frameworks include release date, core language, user-facing API, computation model, communication model, data parallelism, model parallelism, programming paradigm, fault tolerance, and UI/visualization. This choice of criteria is explained in Section 2. Tensorflow, Deeplearning4j, MXNet, H2O, and CaffeOnSpark were chosen by a combination of factors, including their being open-source, level of documentation, maturity as a product, and adoption by the community. Frameworks currently without distributed support won't be the focus of this discussion, but well-known ones such as Theano, Torch, Caffe (without Spark) have been studied \cite{DBLP:journals/corr/BahrampourRSS15}.

@@

\section{Framework Comparison}
\begin{tabularx}{1.3\textwidth}{ |X|X|X|X|X|X| }
  \hline
  Platform & Tensorflow & Deeplearning4j & MXNet & H2O & CaffeOnSpark \\
  \hline
  Release Date & 2016 (distributed) & 2015 & 2015 & 2014 (deep learning) & 2016 \\
   \hline
  Core Language & C++  & Java  & C++ & Java & \pbox{20cm}{C++ (Caffe) \\ Scala (Spark)}  \\
  \hline
  API & C++, Python & Java, Scala & C++, Python, R, Scala, Matlab, Javascript, Go, Julia & Java, R, Python, Scala, Javascript, web-UI & Python, Matlab, Scala \\
  \hline
  Computation Model & Sync or async & Sync & Sync or async & Async & Sync \\
  \hline
  Communication Model & Parameter server & Iterative MapReduce & Parameter server & Distributed fork-join & MPI Allreduce \\
  \hline
  Data Parallelism & \checkmark & \checkmark & \checkmark& \checkmark & \checkmark \\
  \hline
  Model Parallelism & \checkmark & \xmark & \checkmark & \xmark & \xmark \\
  \hline
  Programming Paradigm & Imperative & Declarative & Both & Declarative & Declarative \\
  \hline
  Fault Tolerance & Checkpoint-and-recovery & Checkpoint-and-resume & Checkpoint-and-resume & N/A & N/A \\
  \hline
  UI/Visualization & Graph-visualization, training monitoring & Training monitoring & None & None & Summary Statistics \\
  \hline
\end{tabularx}

\section{Framework Discussion}
\subsection{MXNet}
MXNet is a distributed deep learning framework that became available in 2015. It was developed with collaborators from several institutions, including CMU, University of Washington, and Microsoft. It currently interfaces with C++, Python, R, Scala, Matlab, Javascript, Go, and Julia. MXNet supports both declarative and declarative expressions; symbolic in declaring computation graphs with higher-level abstractions like convolutional layers, and imperative in the ability to direct tensor computation and control flow [1]. Data parallelism is supported by default, and it also seems possible to build with model parallelism. Distributed execution in MXNet generally follows a parameter server model, with parallelism and data consistency managed at two levels: intra-worker and inter-worker. Devices within a single worker machine maintain synchronous consistency on its parameters. Inter-worker data consistency can either be synchronous, where gradients over all workers are aggregated before proceeding, or asynchronous, where each worker independently updates parameters. This trade-off between performance and convergence speed is left as an option to the user. The actual handling of server updates and requests is pushed down to MXNet's dependency engine, which schedules all operations and performs resource management. Fault tolerance on MXNet involves checkpoint-and-resume, which must be user-initiated.

@@

>>>>

The goal of this paper is to survey the landscape of deep learning frameworks with support for distributed execution, and compare them according to a consistent set of characteristics. The characteristics compared across frameworks include release date, core language, user-facing API, computation model, communication model, data parallelism, model parallelism, programming paradigm, fault tolerance, and visualization. This choice of criteria is explained in Section 2. Tensorflow, Deeplearning4j, MXNet, H2O, and CaffeOnSpark were chosen by a combination of factors, including their being open-source, level of documentation, maturity as a product, and adoption by the community. Frameworks currently without distributed support won't be the focus of this discussion, but well-known ones such as Theano, Torch, Caffe (without Spark) have been studied \cite{DBLP:journals/corr/BahrampourRSS15}.

@@

\section{Framework Comparison}
\begin{tabularx}{1.3\textwidth}{ |X|X|X|X|X|X| }
  \hline
  Platform & Tensorflow & Deeplearning4j & MXNet & H2O & CaffeOnSpark \\
  \hline
  Release Date & 2016 (distributed) & 2015 & 2015 & 2014 (deep learning) & 2016 \\
   \hline
  Core Language & C++  & Java  & C++ & Java & \pbox{20cm}{C++ (Caffe) \\ Scala (Spark)}  \\
  \hline
  API & C++, Python & Java, Scala & C++, Python, R, Scala, Matlab, Javascript, Go, Julia & Java, R, Python, Scala, Javascript, web-UI & Python, Matlab, Scala \\
  \hline
  Computation Model & Sync or async & Sync & Sync or async & Async & Sync \\
  \hline
  Communication Model & Parameter server & Iterative MapReduce & Parameter server & Distributed fork-join & MPI Allreduce \\
  \hline
  Data Parallelism & \checkmark & \checkmark & \checkmark& \checkmark & \checkmark \\
  \hline
  Model Parallelism & \checkmark & \xmark & \checkmark & \xmark & \xmark \\
  \hline
  Programming Paradigm & Imperative & Declarative & Both & Declarative & Declarative \\
  \hline
  Fault Tolerance & Checkpoint-and-recovery & Checkpoint-and-resume & Checkpoint-and-resume & N/A & N/A \\
  \hline
  Visualization & Graph-visualization, training monitoring & Training monitoring & None & None & Summary Statistics \\
  \hline
\end{tabularx}

The relevance of release date, core language, user-facing APIs are self-explanatory. Computation model specifies the nature of data consistency through execution, i.e. whether updates are synchronous or asynchronous. In context of optimization kernels like stochastic gradient descent (SGD), synchronous execution has better convergence guarantees by maintaining consistency or near-consistency with sequential execution. However, asynchronous SGD can exploit more parallelism and train faster, but with less guarantees of convergence speed. Note that frameworks like Tensorflow and MXNet simply leave this as a tradeoff choice for the user.

The communication model tries to categorize the nature of distributed execution by well-known paradigms.

Data and model parallelism are the two prevalent opportunities for parallelism in training deep learning networks at the distributed level. In data parallelism, copies of the model, or parameters, are each trained on its own subset of the training data, while updating the same global model. In model parallelism, the model itself is partitioned and trained in parallel.

Programming paradigm falls into the categories of imperative, declarative, or a mix of both. Conventionally, imperative programming specifies \textit{how} a computation is done, where as declarative programming specifies \textit{what} needs to be done. There is plenty of gray area, but the distinction is made in this paper based on whether the API exposes the user to computation details that require some understanding of the inner math of neural networks (imperative), or whether the abstraction is yet higher (declarative).

Fault tolerance is included for two reasons. Distributed execution tends to be more failure prone, especially at scale. Furthermore, any failures (not necessarily limited to distributed execution) that interrupt training part-way can be very costly, if all the progress made on the model is simply lost.

Finally, UI/Visualization is a feature supported to very different degrees across the frameworks studied. The ability to monitor the progress of training and the internal state of networks over time could be useful for debugging or hyperparameter tuning, and could be an interesting direction. Tensorflow and Deeplearning4j both support this kind of visualization.

The scope of deep learning functionality was a possible point of discussion. However, initial surveying suggested that at this point most deep learning libraries provide all the standard network architectures, training and optimization techniques. More differences are likely to be found in programmability or performance.

\section{Framework Discussion}

@@

\subsection{MXNet}
MXNet is a distributed deep learning framework that became available in 2015. It was developed in collaboration across several institutions, including CMU, University of Washington, and Microsoft. It currently interfaces with C++, Python, R, Scala, Matlab, Javascript, Go, and Julia. MXNet supports both declarative and declarative expressions; symbolic in declaring computation graphs with higher-level abstractions like convolutional layers, and imperative in the ability to direct tensor computation and control flow [1]. Data parallelism is supported by default, and it also seems possible to build with model parallelism. Distributed execution in MXNet generally follows a parameter server model, with parallelism and data consistency managed at two levels: intra-worker and inter-worker. Devices within a single worker machine maintain synchronous consistency on its parameters. Inter-worker data consistency can either be synchronous, where gradients over all workers are aggregated before proceeding, or asynchronous, where each worker independently updates parameters. This trade-off between performance and convergence speed is left as an option to the user. The actual handling of server updates and requests is pushed down to MXNet's dependency engine, which schedules all operations and performs resource management. Fault tolerance on MXNet involves checkpoint-and-resume, which must be user-initiated.
